---
title: "carbayes_modelling"
output: html_document
---

```{r import-libraries-installs}
# install.packages('CARbayes')
library(CARBayes)
# install.packages('CARBayesST')
library(CARBayesST)
library(tidyverse)
# install.packages("BayesianTools")
library(BayesianTools)
```


Rescale: populationdensity/10
```{r import-data (insert file location)}
data <- read.csv('clean_data/fireClimateHumanTable_editted.csv')
data$populationdensity <- data$populationdensity/10
```


Building Neighborhood Matrix
```{r (insert file location)}
neighborhood_matrix <- read.csv('clean_data/neighborhoodMatrix.csv', header=FALSE)
neighborhood_matrix <- as.matrix(neighborhood_matrix)
W <- matrix(NA, 56, 56)
for(i in 1:56){
  for(j in 1:56){
    W[i,j] = as.numeric(neighborhood_matrix[i+1,j+1])
  }
}
```

Our final model:
```{r month_factor_model: count ~ X2m_temperature + low_vegetation_cover + X10m_wind_speed + populationdensity}
final_model <- ST.CARlinear(formula = count ~ X2m_temperature + low_vegetation_cover + X10m_wind_speed + populationdensity,
              family = 'poisson',
              data = data,
              W = W,
              burnin = 5000,
              n.sample = 10000)
```

Here we are looking at the summary of the final model, as well as its coeffecients (these are median coefficients)

```{r month_factor_model results}
final_model
coef(final_model)  
```

This gives us more information about the carlinear model

```{r names() of model}
names(final_model)
final_model$formula
```

Here we are summarizing the posterior distribution of the temperature variable
We retrieve it by indexing on samples\$beta, and then the column of choice (example: vegetation would be samples$beta[,2])
```{r posterior distribution of temperature}
hist(final_model$samples$beta[,2], xlab = "coeff of temperature", main = "posterior distribution of temperature") 
```

More visualizations:
```{r traceplot of X10m_wind_speed}
plot(final_model$samples$beta[,2], main= "X2m_temperature DensityPlot")
hist(final_model$samples$beta[,2], xlab = "coeff of X2m_temperature", main = "Posterior Distribution of X2m_temperature") 
hist(final_model$samples$beta[,4], xlab = "coeff of X10m_windspeed", main = "Posterior Distribution of X10m_wind_speed") 
```









