---
title: "CarbayesMSE"
output: html_document
---

```{r}
#install.packages('CARbayes')
#install.packages('CARBayesST')
library(CARBayes)
library(tidyverse)
library(CARBayesST)
library('lme4')
library(tidyverse)
library(janitor)
library(ggplot2)
library(dplyr)
```

Load the data from the clean data folder: INSERT PATH HERE, and scale population density
```{r loading data}
data <- read.csv('clean_data/fireClimateHumanTable_editted.csv')
data$populationdensity <- data$populationdensity/10
```
Load the neighborhood matrix and format it

```{r load neighborhood matrix}
neighborhood_matrix <- read.csv('clean_data/neighborhoodMatrix.csv', header=FALSE)
neighborhood_matrix <- as.matrix(neighborhood_matrix)
W <- matrix(NA, 56, 56)
for(i in 1:56){
  for(j in 1:56){
    W[i,j] = as.numeric(neighborhood_matrix[i+1,j+1])
  }
}
```

FOR CARBAYES:

The model:

```{r}
model <- ST.CARlinear(formula = "count ~ X2m_temperature + low_vegetation_cover + X10m_wind_speed + populationdensity",
            family = 'poisson',
            data = data,
            W = W,
            burnin = 5000,
            n.sample = 10000)
```

The MSE function:

```{r}
train_test_validation <- function(formula_string, data, W){
  total_MSE = 0
  for (y in 2013:2019){
    train<- data %>%
    subset(year!=y)
    test<-data %>%
    subset(year==y)
  
    model <- ST.CARlinear(formula = formula_string,
                family = 'poisson',
                data = train,
                W = W,
                burnin = 500,
                n.sample = 1000)
    total = matrix(NA,  nrow = dim(test)[1], ncol = dim(model$samples$beta)[1] )
    samples.beta = as.matrix(model$samples$beta)
    samples.phi = as.matrix(model$samples$phi)
    samples.delta = as.matrix(model$samples$delta)
    samples.alpha = as.vector(model$samples$alpha[,1])
    t = as.vector(rep(73:84, each=56))
    county <- rep(1:56,12)
    N = length(train$count)/56
    t.bar = sum(1:72)/72
    total.median = rep(NA, each=dim(test)[1])
    for(i in 1:dim(test)[1]){
      for(j in 1:dim(model$samples$beta)[1])
      {
        total[i,j] =as.numeric(exp(sum(samples.beta[j,]*c(1,test$X2m_temperature[i],
                                                          test$low_vegetation_cover[i],
                                                          test$X10m_wind_speed[i],
                                                          test$populationdensity[i])) + samples.phi[j,test$county_factor[i]] + (samples.alpha[j] + samples.delta[j,test$county_factor[i]])*((test$month_and_year_factor[i] - t.bar)/72)))
      }
      total.median[i] = median(total[i,])
    }
    
    mse = 0
    for(i in 1:length(total.median)){
      mse = mse + (test$count[i] - total.median[i])**2
    }
    mse = mse/length(total.median)
    total_MSE = total_MSE + mse
    
  }
  return(total_MSE/7)
}
```


```{r}
train_test_validation("count ~ X2m_temperature + low_vegetation_cover + X10m_wind_speed + populationdensity", data, W)
```

Forecasting:

Predict one data row:
```{r}
forecast_one_row <- function(model, data, i){
  total = rep(NA, dim(model$samples$beta)[1] )
  samples.beta = as.matrix(model$samples$beta)
  samples.phi = as.matrix(model$samples$phi)
  samples.delta = as.matrix(model$samples$delta)
  samples.alpha = as.vector(model$samples$alpha[,1])
  t.bar = sum(1:72)/72
  for(j in 1:dim(model$samples$beta)[1])
    {
      total[j] =as.numeric(exp(sum(samples.beta[j,]*c(1,data$X2m_temperature[i],
                                                        data$low_vegetation_cover[i],
                                                        data$X10m_wind_speed[i],
                                                        data$populationdensity[i])) + samples.phi[j,data$county_factor[i]] + (samples.alpha[j] + samples.delta[j,data$county_factor[i]])*((data$month_and_year_factor[i] - t.bar)/72)))
  }
  
  return (median(total))
}
```

Predict one county:
```{r}
forecast_one_county <- function(model, data, county_str){
  county_data = data %>% 
    subset(data$county == county_str)
  total = matrix(NA,  nrow = dim(county_data)[1], ncol = dim(model$samples$beta)[1] )
  samples.beta = as.matrix(model$samples$beta)
  samples.phi = as.matrix(model$samples$phi)
  samples.delta = as.matrix(model$samples$delta)
  samples.alpha = as.vector(model$samples$alpha[,1])
  #county <- rep(1:56,12)
  #N = length(train$count)/56
  t.bar = sum(1:72)/72
  #total.median = rep(NA, each=dim(test)[1])
  total.median = rep(NA, each=dim(county_data)[1])
  for(i in 1:dim(county_data)[1]){
    for(j in 1:dim(model$samples$beta)[1])
      {
        total[i,j] =as.numeric(exp(sum(samples.beta[j,]*c(1,county_data$X2m_temperature[i],
                                                          county_data$low_vegetation_cover[i],
                                                          county_data$X10m_wind_speed[i],
                                                          county_data$populationdensity[i])) + samples.phi[j,county_data$county_factor[i]] + (samples.alpha[j] + samples.delta[j,county_data$county_factor[i]])*((county_data$month_and_year_factor[i] - t.bar)/72)))
    }
  total.median[i] = median(total[i,])
  }
  county_data["predicted_count"] = total.median
  
  return (county_data)
}
```


Predict one month:
```{r}
forecast_one_month <- function(model, data, month_number, year_number){
  month_data = data %>% 
    subset(data$month_factor == month_number & data$year == year_number)
  total = matrix(NA,  nrow = dim(month_data)[1], ncol = dim(model$samples$beta)[1] )
  samples.beta = as.matrix(model$samples$beta)
  samples.phi = as.matrix(model$samples$phi)
  samples.delta = as.matrix(model$samples$delta)
  samples.alpha = as.vector(model$samples$alpha[,1])
  t.bar = sum(1:72)/72
  total.median = rep(NA, each=dim(month_data)[1])
  for(i in 1:dim(month_data)[1]){
    for(j in 1:dim(model$samples$beta)[1])
      {
        total[i,j] =as.numeric(exp(sum(samples.beta[j,]*c(1,month_data$X2m_temperature[i],
                                                          month_data$low_vegetation_cover[i],
                                                          month_data$X10m_wind_speed[i],
                                                          month_data$populationdensity[i])) + samples.phi[j,month_data$county_factor[i]] + (samples.alpha[j] + samples.delta[j,month_data$county_factor[i]])*((month_data$month_and_year_factor[i] - t.bar)/72)))
    }
  total.median[i] = median(total[i,])
  }
  month_data["predicted_count"] = total.median
  
  return (month_data)
}
```

One sample forecasting:
```{r}
model <- ST.CARlinear(formula = "count ~ X2m_temperature + low_vegetation_cover + X10m_wind_speed + populationdensity",
            family = 'poisson',
            data = data,
            W = W,
            burnin = 5000,
            n.sample = 10000)
month_forecast <- forecast_one_month(model, data, 6, 2019)
```

Posterior predictive data for visualizing
```{r}
posterior_predictive_check <- function(model, data, county_str){
  county_data = data %>% 
    subset(data$county == county_str & year == 2019)
  total = matrix(NA,  nrow = dim(county_data)[1], ncol = dim(model$samples$beta)[1] )
  fake_data = matrix(NA,  nrow = dim(county_data)[1], ncol = dim(model$samples$beta)[1] )
  samples.beta = as.matrix(model$samples$beta)
  samples.phi = as.matrix(model$samples$phi)
  samples.delta = as.matrix(model$samples$delta)
  samples.alpha = as.vector(model$samples$alpha[,1])
  t.bar = sum(1:72)/72
  total.median = rep(NA, each=dim(county_data)[1])
  for(i in 1:dim(county_data)[1]){
    for(j in 1:dim(model$samples$beta)[1])
      {
        total[i,j] =as.numeric(exp(sum(samples.beta[j,]*c(1,county_data$X2m_temperature[i],
                                                          county_data$low_vegetation_cover[i],
                                                          county_data$X10m_wind_speed[i],
                                                          county_data$populationdensity[i])) + samples.phi[j,county_data$county_factor[i]] + (samples.alpha[j] + samples.delta[j,county_data$county_factor[i]])*((county_data$month_and_year_factor[i] - t.bar)/72)))
        fake_data[i,j] = rpois(1, total[i,j])
    }
  total.median[i] = median(total[i,])
  }
  total = data.frame(total)
  fake_data = data.frame(fake_data)
  total$month= county_data$month_factor
  fake_data["month"] = county_data$month_factor
  transpose_total = t(total)
  transpose_fake_data = t(fake_data)
  write.csv(total,"totalCarBayes.csv")
  write.csv(fake_data,"fake_data.csv")
  write.csv(transpose_total,"transpose_totalCarBayes.csv")
  write.csv(transpose_fake_data,"transpose_fake_data.csv")
  
  
  
  stripchart(transpose_total)
  

  
  
  
  return (total, fake_data)
}
```


MIXED EFFECTS:

Scaling the data
```{r}
temp_data <- scale(data[,10:19])
data[,10:19] <- temp_data
```


The model:
```{r}
finalized_model <- glmer(formula=count ~ X2m_temperature + low_vegetation_cover + X10m_wind_speed  + populationdensity + (1|county) + (1|month_factor), data = data, family='poisson')
```


calculate MSE:
```{r}
total_mse <- 0
for (y in 2013:2019){
    train<- data %>%
      subset(year!=y)
    test<-data %>%
      subset(year==y)
    model <- glmer(formula=count ~ X2m_temperature + low_vegetation_cover + X10m_wind_speed  + populationdensity + (1|county) + (1|month_factor), data = train, family='poisson')
    pred <- predict(model, test)
   mse <- 0
   preds <- as.vector(pred)
   for(i in 1:length(test)){
      mse = mse + ((test$count[i] - exp(preds[i]))**2)
   }
   total_mse = total_mse + mse/length(test)
}
total_mse/7
```

forecasting for a specific county, across all months
```{r}
county_data = data %>% 
    subset(data$county == 'Riverside' & data$year == 2019)
pred_riverside <- as.vector(exp(predict(model, county_data)))
county_data["predicted_count"] <- pred_riverside
```


forecasting for a specific month, across all counties
```{r}
month_data = data %>% 
    subset(month_factor == 6 & year == 2019)
pred_july <- as.vector(exp(predict(model, month_data)))
month_data["predicted_count"] <- pred_july
```
















